Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-17 forecastflows.tex
adding some stuff about pbmodels (but not great)

\begin{figure}[h]
	\centering
	\includegraphics{figures/modeldataspaces}
	\caption{Here the caption could explain a bit more the pre-model pre-data to the post-model post-data...} 
	\label{fig:modeldata}
\end{figure}
% jdavies15april: Could the bottom right be where exploratory data analysis fits in, generating hypotheses that inform the research question?

@@

Applying the workflow to process-based models would help open the black box.
Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two and help expose potential degeneracies in the model design. %jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below) [FLAG: agree, and made me realize it was not 100% clear in my mind. As I understand it, non-identifiability concerns one parameter, and degeneracy concerns the all model?... but not sure]
Uncovering degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight degeneracies by requiring more explicit model calibration (i.e. data fitting), % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated). Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections. 
% jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties? This could help illustrate potential degeneracies etc.

% jdavies15april: might be better grouped together in concluding section
Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.

>>>>

\begin{figure}[h]
	\centering
	\includegraphics{figures/modeldataspaces_rotation}
	\caption{Here the caption could explain a bit more the pre-model pre-data to the post-model post-data...} 
	\label{fig:modeldata}
\end{figure}
% jdavies15april: Could the bottom right be where exploratory data analysis fits in, generating hypotheses that inform the research question?

@@

% jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties? This could help illustrate potential degeneracies etc.
% [FLAG: I tried something, but I'm not really illustrating degeneracies + I don't like citing myself as much + I'm not really satisfied]
An illustration of the issues with the current forecast workflow is the process-based models used to project species or ecosystem distributions. Most of the focus is on the model projections---often represented as maps without uncertainties. Model evaluation generally focuses on how well these projections match observed distributions, sometimes under different climatic conditions to challenge the models \citep{VanderMeersch2025a}---most of the time evaluating only of the many different output variables of the model. The model building and calibration steps are usually limited to the Methods section, and estimated parameter values are rarely discussed \citep[e.g.][]{Conradi2024, VanderMeersch2025a}. The complexity of these models often makes it difficult to assess the robustness of their calibration, whether for a reviewer or a reader. If ever done, the assesment of the model degeneracies is often done as a a separate work, exposing model degeneracies later on \citep{VanderMeersch2025b}, and not required as a preliminary step.

Applying the workflow to process-based models would help open the black box.
Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting (also called calibration), ensuring a clear delineation between the two and help expose potential degeneracies in the model design. %jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below) [FLAG: agree, and made me realize it was not 100% clear in my mind. As I understand it, non-identifiability concerns one parameter, and degeneracy concerns the all model?... but not sure]
Uncovering degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight degeneracies by requiring more explicit model calibration (i.e. data fitting), % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated). Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.

% jdavies15april: might be better grouped together in concluding section
% [FLAG: I feel if we want to put this in the concluding section, we would need to talk also about the "opportunities to bridge statistical and process-based", rather than keeping it in the box? So, what should we put in the box?]
Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.
